GSEA datasets for FDR benchmarking

### 2018-02-22 datasets/GSEA/gsea.Rmd
Standardize format of gsea

---
title: "GSEA datasets for FDR benchmarking"
author: "Alejandro Reyes (on behalf of the Rafalab Benchmarking Consortium)"
output: html_document
---

@@

# Differential expression analysis with DESeq2

Downloading *DESeqDataSet* if not present locally,

@@

file <- "dsd.RData"
if(!file.exists( file ) ){
    download.file("https://www.dropbox.com/s/rdmv3tyqt4g37b0/dsd.RData?dl=1",destfile="dsd.RData")
}
load(file)

library(DESeq2)
levels(colData(dsdObject)$individual) <- gsub("GTEX-", "",levels(colData(dsdObject)$individual))

@@

Specify model matrix design and run DESeq2 to get differentially expressed genes.

```{r deseq2, message=FALSE}

library(EnsDb.Hsapiens.v86)
gnType <-  genes(EnsDb.Hsapiens.v86, column="gene_biotype")

dsdObject <- dsdObject[rownames(dsdObject) %in% gnType[gnType$gene_biotype == "protein_coding"]$gene_id,]
dsdObject <- DESeq(dsdObject)
res <- results( dsdObject, independentFiltering=FALSE )

@@

Performing a gene set enrichment analysis with goseq.

```{r goseq, message=FALSE}

library(goseq)

file="goSets.RData"
if(!file.exists(file)){
  stopifnot( require( biomaRt ) )
  mart <- useMart( "ensembl", "hsapiens_gene_ensembl" )
  goSets <- getBM( c("ensembl_gene_id", "go_id"), mart=mart, filters="ensembl_gene_id", values=rownames(res))
  goSets <- goSets[!nchar( goSets$go_id ) == 0,]
  goSets <- with( goSets, split( go_id, ensembl_gene_id ) )
  save(goSets, file=file)
}else{
  load(file)
}

genes <- as.numeric( res$padj < 0.1 )
names( genes ) <- rownames( res )

```{r informativeCovariate, message=FALSE, fig.height=3.5, message=FALSE, fig.width=4.5}

library(ggplot2)
library(magrittr)
library(cowplot)
library(R.utils)
library(scales)
sourceDirectory("../R/")

```{r assumptions, message=FALSE, fig.width=10, fig.height=3.2}

library(dplyr)

## filter numInCat below only used for plotting
## filter numDEInCat for cases where there are no DE genes in the category (these are actually not tested)
#dplyr:::filter( goRes, over_represented_pvalue == 1, numDEInCat > 0)

@@

@@

# Gene set size of overlaps

@@

# NULL comparisons

@@

### preparing gene length info and goSets ###
library("EnsDb.Mmusculus.v75")
txByGene <- transcriptsBy(EnsDb.Mmusculus.v75, "gene")
geneLength <- sapply( width( txByGene ), median )
geneLength <- geneLength[names(genes)]

@@

library(dplyr)

>>>>

---
title: "GSEA datasets for FDR benchmarking"
author: "Alejandro Reyes (on behalf of the Rafalab Benchmarking Consortium)"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
    html_document:
        toc: true
        toc_float: true
        highlight: tango
        number_sections: true
---

@@

# Summary

Here we will make use of the `SummarizedBenchmark` package to benchmark two 
Gene Set Enrichment Analysis (GSEA) case studies.

# Workspace setup

```{r, wkspace-setup, results='hide', message=FALSE, warning=FALSE}
# data and results directories
system("mkdir -p DATA")
system("mkdir -p RESULTS")
datdir <- "DATA/"
resdir <- "RESULTS/"

# GSEA analysis tools
library(DESeq2)
library(goseq)
library(EnsDb.Hsapiens.v86)
library(EnsDb.Mmusculus.v75)
library(biomaRt)

# wrangling & plotting tools
library(ggplot2)
library(cowplot)
library(scales)
library(dplyr)

# comparison tools/functions
library(R.utils)
sourceDirectory("../R/")
```

# GTEX expression data

## Differential expression analysis with DESeq2

First we'll download the dataset from GTEX, which is saved as an `RData` file and
contains a *DESeqDataSet* object.

@@

file <- paste0(datdir, "dsd.RData")
if(!file.exists( file ) ){
    download.file("https://www.dropbox.com/s/rdmv3tyqt4g37b0/dsd.RData?dl=1",
                  destfile=file)
}
load(file)

levels(colData(dsdObject)$individual) <- gsub("GTEX-", "",levels(colData(dsdObject)$individual))

@@

Specify model matrix design and run **DESeq2** to get differentially expressed genes.

```{r deseq2, message=FALSE}
gnType <-  genes(EnsDb.Hsapiens.v86, column="gene_biotype")

dsdObject <- dsdObject[rownames(dsdObject) %in% gnType[gnType$gene_biotype == "protein_coding"]$gene_id,]
dsdObject <- DESeq(dsdObject)
res <- results( dsdObject, independentFiltering=FALSE )
```

## Gene set enrichment analysis with goseq

Next we'll use the 
**biomaRt** package to download GO gene set annotations (for humans since the
GTEX data contains human samples).

```{r biomart, message=FALSE}
file <- paste0(datdir, "goSets.RData")
if(!file.exists(file)){
  mart <- useMart( "ensembl", "hsapiens_gene_ensembl" )
  goSets <- getBM( c("ensembl_gene_id", "go_id"), mart=mart, 
                   filters="ensembl_gene_id", values=rownames(res))
  goSets <- goSets[!nchar( goSets$go_id ) == 0,]
  goSets <- with( goSets, split( go_id, ensembl_gene_id ) )
  save(goSets, file=file)
}else{
  load(file)
}
```

Finally, we'll perform a gene set enrichment analysis with **goseq**. We'll 
include all genes with a **DESeq2** adjusted p-value less than 0.10, and adjust
for gene length since not doing so may bias the results toward GO sets with 
longer genes.

```{r goseq, message=FALSE}
genes <- as.numeric( res$padj < 0.1 )
names( genes ) <- rownames( res )

pwf <- nullp(genes, bias.data=geneLength )
goRes <- goseq( pwf, gene2cat=goSets )
```

## Check Covariate Diagnostics

```{r informativeCovariate, message=FALSE, fig.height=3.5, message=FALSE, fig.width=4.5}
rank_scatter( dat=goRes, pval="over_represented_pvalue", 
              covariate="numInCat", bins=50, funx=log2, 
              funfill=log10_trans()) +
    ylim(0, 12) + ggtitle("Over-represented gene sets") +
    xlab( expression(log[10]~"(# of genes)")) +
    ylab( expression(-log[10]~"(p-value)") )

@@

```{r assumptions, message=FALSE, fig.width=10, fig.height=3.2}
## filter numInCat below only used for plotting
## filter numDEInCat for cases where there are no DE genes in the category (these are actually not tested)
#dplyr:::filter( goRes, over_represented_pvalue == 1, numDEInCat > 0)

@@

## Run benchmarks on set size covariate

@@

## Gene set size of overlaps

@@

## NULL comparisons

@@

### preparing gene length info and goSets ###
txByGene <- transcriptsBy(EnsDb.Mmusculus.v75, "gene")
geneLength <- sapply( width( txByGene ), median )
geneLength <- geneLength[names(genes)]

@@
